---
title: "Analysis of Missing Values in NDA Dataset"
date: "`r Sys.Date()`"
author: Siti Hassan
date-format: "D MMMM YYYY"
format:
  html:
    # Table of Contents options
    toc: true
    toc-depth: 3
    toc-expand: true
    toc-location: left
    toc-title: Table of Contents
    number-sections: false
    number-depth: 3
    # Render options
    theme: cosmo
    css: CSS/theme.css   # Replace with the directory of your css file (e.g., theme.css within 'CSS' folder)
    anchor-sections: false
    html-math-method: katex
    # Code options
    code-tools:
      source: false
      toggle: false
      caption: none
    code-fold: false
    # code-summary: "Show code"
    embed-resources: true
    standalone: true
    # URL options
    link-external-icon: false
    link-external-newwindow: true
    # Reference options
    citations-hover: true
    footnotes-hover: true
    # Callout options
    callout-appearance: simple
    callout-collapse: true
    # Caption options
    cap-location: bottom
    # Title options
    title-block-banner: '#005EB8'
    backgroundcolor: '#f0f4f5'
    # Set font
    mainfont: 'Open Sans'
    fig-width: 7
    fig-height: 6
      
execute:
  echo: false
  cache: true              # Cache results
  warning: false           # Do not include warnings in rendered output
  error: false             # Do not include errors in the output
  feeze: auto              # Re-compute previously generated computational output only in case their source file changes
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
include-after-body: "CSS/footer.html"
---

```{r}
#| label: load-libraries
#| output: false


# Load the required library
library(ggplot2)
library(plotly)
library(tidyverse)
library(naniar)
library(DT)

```

# Dataset

```{r}
#| label: load-data
#| output: false

# Read the RDS data
data <- readRDS("../data/nda_clean.rds")
```

The National Diabetes Audit (NDA) [dataset](https://digital.nhs.uk/data-and-information/clinical-audits-and-registries/national-diabetes-audit) contains data on care processes and treatment targets for diabetes care, enabling us to measure the effectiveness of diabetes healthcare against NICE Clinical Guidelines and NICE Quality Standards in England and Wales. 

This report examines the quality of the NDA dataset, focusing specifically on the distribution of missing values within it. This represents the first step in understanding the dataset before utilizing it for any modelling work. 

The dataset provides a comprehensive audit of diabetes care and outcomes for patients in the Birmingham and Solihull, extracted for the 2023-24 financial year (2023-24 Q4). 



```{r}
#| label: plotting-functions

create_lollipop_plot <- function(data, x_var, y_var, plot_title, xaxis_title, yaxis_title, showlegend = FALSE, horizontal = TRUE, marker_color, line_color) {
  
  # Convert the arguments to tidy column names
  x_var <- rlang::as_name(rlang::ensym(x_var))
  y_var <- rlang::as_name(rlang::ensym(y_var))
  
  # Dynamically process data based on orientation
  if (horizontal) {
    # Sort by x_var (the horizontal variable) and reverse factor levels for y_var
    processed_data <- data %>%
      arrange(desc(.data[[x_var]])) %>%
      mutate(!!y_var := factor(.data[[y_var]], levels = rev(.data[[y_var]])))
  } else {
    # Sort by y_var (the vertical variable) and reverse factor levels for x_var
    processed_data <- data %>%
      arrange(desc(.data[[y_var]])) %>%
      mutate(!!x_var := factor(.data[[x_var]], levels = rev(.data[[x_var]])))
  }
  
  # Create the lollipop-style chart
  interactive_plot <- plot_ly() %>%
    # Add the segments (lines)
    add_segments(
      data = processed_data,
      x = if (horizontal) ~0 else ~.data[[x_var]], 
      xend = if (horizontal) ~.data[[x_var]] else ~.data[[x_var]],
      y = if (horizontal) ~.data[[y_var]] else ~0, 
      yend = if (horizontal) ~.data[[y_var]] else ~.data[[y_var]],
      line = list(color = line_color),
      hoverinfo = 'none' # No hover info for lines
    ) %>%
    # Add the points
    add_markers(
      data = processed_data,
      x = if (horizontal) ~.data[[x_var]] else ~.data[[x_var]],
      y = if (horizontal) ~.data[[y_var]] else ~.data[[y_var]],
      marker = list(size = 8, color = marker_color), # Customize marker
      hoverinfo = 'x+y' # Show x and y variables on hover
    ) %>%
    # Customize layout
    layout(
      title = plot_title,
      xaxis = list(title = xaxis_title),
      yaxis = list(title = yaxis_title),
      margin = list(l = 150), # Add space for long variable names
      showlegend = showlegend # Show/hide the legend based on input
    )
  
  # Return the plot
  return(interactive_plot)
}

```
# Missing Values
## Missingness in Cases

This plot shows the number and percentage of missing values for each case. 


:::: panel-tabset

## # Missing
```{r}

# Summarize missing cases and group by n_miss
missing_dt <- miss_case_summary(data) %>%
  group_by(n_miss) %>%
  summarise(count = n()) %>%
  arrange(n_miss) # Ensure ascending order of n_miss

# Create the bar chart
p1 <- plot_ly(
  data = missing_dt,
  x = ~n_miss,       # Number of missing values
  y = ~count,        # Count of cases
  type = "bar",      # Bar chart
  marker = list(color = "#45837FFF", opacity = 0.8) # Customize marker
) %>%
  layout(
    title = "Distribution of Missing Data per Case",
    xaxis = list(title = "# Missing"),   # Set x-axis title
    yaxis = list(title = "Cases"),       # Set y-axis title
    bargap = 0.1                         # Adjust gap between bars
  )

# Display the plot
p1
```
## % Missing

```{r}

# Summarize missing cases and group by n_miss
missing_dt <- miss_case_summary(data) %>%
  group_by(n_miss) %>%
  summarise(count = n()) %>%
  mutate(total = sum(count),
         pct_miss = (count/total) * 100,
         pct_miss = round(pct_miss, 2))

# Create the bar chart
p1 <- plot_ly(
  data = missing_dt,
  x = ~n_miss,       # Number of missing values
  y = ~pct_miss,        # Count of cases
  type = "bar",      # Bar chart
  marker = list(color = "#E5AD4FFF", opacity = 0.8) # Customize marker
) %>%
  layout(
    title = "Distribution of Missing Data per Case",
    xaxis = list(title = "# Missing"),   # Set x-axis title
    yaxis = list(title = "Cases"),       # Set y-axis title
    bargap = 0.1                         # Adjust gap between bars
  )

# Display the plot
p1

```

## Table


```{r}
# Summarize missing cases and group by n_miss
missing_dt <- miss_case_summary(data) %>%
  group_by(n_miss) %>%
  summarise(count = n()) %>%
  mutate(total = sum(count),
         pct_miss = (count/total) * 100,
         pct_miss = round(pct_miss, 2)) %>% 
  mutate(count = format(count, big.mark = ",")) %>% 
  select(n_miss, count, pct_miss)

datatable(missing_dt, extensions = "Buttons", 
            options = list(paging = TRUE,
                           scrollX=TRUE, 
                           searching = TRUE,
                           ordering = TRUE,
                           dom = 'Bfrtip',
                           buttons = c('copy', 'csv', 'excel', 'pdf'),
                           pageLength=5, 
                           lengthMenu=c(3,5,10) ))

```

::::

::: {.callout-tip}

# Summaries

- There are **48,065 cases or rows (42.49%)** with no missing values.

- There are **27,286 cases (13.22%)** with exactly 2 missing values.

- The number of cases with missing values progressively decreases as the number of missing values increases.

:::

## Missingness in Variables

This plot shows the number and percentage of missing values for each variable in the dataset. 

:::: panel-tabset

## # Missing
```{r}
#| label: plot-missingness

# Create the missing data summary
missing_dt <- miss_var_summary(data) %>%
  mutate(
    pct_miss = as.numeric(format(pct_miss, nsmall = 2))) # Ensure numeric for plotting


# Create the plot
create_lollipop_plot(
  data = missing_dt,
  x_var = n_miss,
  y_var = variable,
  plot_title = "Number of Missing Values",
  xaxis_title = " ",
  yaxis_title = " ",
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "#009999",
  line_color = "grey"
)

```
## % Missing

```{r}

# Create the plot
create_lollipop_plot(
  data = missing_dt,
  x_var = pct_miss,
  y_var = variable,
  plot_title = "Percentage of Missing Values",
  xaxis_title = " ",
  yaxis_title = " ",
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)



```

## Table
```{r}
#| label: table-missingness

# Summarize missingness
missing_dt <- missing_dt %>%
  mutate(
    pct_miss = round(as.numeric(pct_miss), 2), # Round to 2 decimal places and keep as numeric
    n_miss = format(n_miss, big.mark = ",")    # Format n_miss with commas
  )

datatable(missing_dt, extensions = "Buttons", 
            options = list(paging = TRUE,
                           scrollX=TRUE, 
                           searching = TRUE,
                           ordering = TRUE,
                           dom = 'Bfrtip',
                           buttons = c('copy', 'csv', 'excel', 'pdf'),
                           pageLength=5, 
                           lengthMenu=c(3,5,10) ))

```

::::

::: {.callout-tip}

# Summaries

- The variables with the highest number of missing values include `clean_cholesterol_value`, the corresponding `cholesterol_category`, and `statin_flag`, each having close to 40,000 missing values. These missing values may not necessary due to data quality issues but rather the absence of the respective treatment care.

- Variables like `bmi_category`, `clean_bmi_value`, and `clean_diastolic_value` also exhibit a substantial number of missing values, but fewer than the top three.

- Variables such as `clean_systolic_value`, `clean_mmol_hba1c_value`, and the corresponding `hba1c_mmol_category` show moderate levels of missingness.

- Several demographic and clinical variables, including `ethnicity`, `smoking_category`, `diabetes_duration_category`, and others, have significantly fewer missing values.

- Variables like `age`, `gender`, and `treatment targets_achieved` have little to no missingness. 


:::

## Missingness in Variables Grouped by Diabetes Type

This plot shows the distribution of missing values across variables grouped by diabetes type (`Type 1`, `Type 2`, and `Other`).

:::: panel-tabset

## # Missing

```{r}

missing_dt <- data %>%
  group_by(diabetes_type) %>%
  miss_var_summary() %>%
  mutate(pct_miss = round(pct_miss, 2)) # Round percentages to 2 decimal places

# Create individual plots for each diabetes type
# Type 1
p1 <- create_lollipop_plot(
  data = missing_dt %>% filter(diabetes_type == "Type 1"),
  x_var = n_miss,
  y_var = variable,
  plot_title = NULL, # Titles will be set in the subplot layout
  xaxis_title = NULL, # Share x-axis title across plots
  yaxis_title = NULL,
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)

# Type 2
p2 <- create_lollipop_plot(
  data = missing_dt %>% filter(diabetes_type == "Type 2"),
  x_var = n_miss,
  y_var = variable,
  plot_title = NULL,
  xaxis_title = NULL,
  yaxis_title = NULL,
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)

# Other
p3 <- create_lollipop_plot(
  data = missing_dt %>% filter(diabetes_type == "Other"),
  x_var = n_miss,
  y_var = variable,
  plot_title = NULL,
  xaxis_title = NULL,
  yaxis_title = NULL,
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)

# Combine the plots into a single faceted plot
facet_plot_combined <- subplot(
  p1, p2, p3,
  nrows = 1,  # Arrange plots in a single row
  shareY = TRUE,
  shareX = TRUE,
  titleX = TRUE,
  titleY = TRUE
) %>%
  layout(
    title = list(
      text = "Number of Missing Values by Diabetes Type",
      y = 1.5  # Move the main title further up
    ),
    annotations = list(
      list(x = 0.16, y = 1.02, text = "Type 1", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14)),
      list(x = 0.50, y = 1.02, text = "Type 2", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14)),
      list(x = 0.83, y = 1.02, text = "Other", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14))
    ),
    xaxis = list(title = " "),  # Set shared x-axis title
    yaxis = list(title = " ")   # Set shared y-axis title
  )

facet_plot_combined

```
## % Missing

```{r}

missing_dt <- data %>%
  group_by(diabetes_type) %>%
  miss_var_summary() %>%
  mutate(pct_miss = round(pct_miss, 2)) # Round percentages to 2 decimal places

# Create individual plots for each diabetes type
# Type 1
p1 <- create_lollipop_plot(
  data = missing_dt %>% filter(diabetes_type == "Type 1"),
  x_var = pct_miss,
  y_var = variable,
  plot_title = NULL, # Titles will be set in the subplot layout
  xaxis_title = NULL, # Share x-axis title across plots
  yaxis_title = NULL,
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)

# Type 2
p2 <- create_lollipop_plot(
  data = missing_dt %>% filter(diabetes_type == "Type 2"),
  x_var = pct_miss,
  y_var = variable,
  plot_title = NULL,
  xaxis_title = NULL,
  yaxis_title = NULL,
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)

# Other
p3 <- create_lollipop_plot(
  data = missing_dt %>% filter(diabetes_type == "Other"),
  x_var = pct_miss,
  y_var = variable,
  plot_title = NULL,
  xaxis_title = NULL,
  yaxis_title = NULL,
  horizontal = TRUE,
  showlegend = FALSE,
  marker_color = "orange",
  line_color = "grey"
)

# Combine the plots into a single faceted plot
facet_plot_combined <- subplot(
  p1, p2, p3,
  nrows = 1,  # Arrange plots in a single row
  shareY = TRUE,
  shareX = TRUE,
  titleX = TRUE,
  titleY = TRUE
) %>%
  layout(
    title = list(
      text = "Percentage of Missing Values by Diabetes Type",
      y = 1.5  # Move the main title further up
    ),
    annotations = list(
      list(x = 0.16, y = 1.02, text = "Type 1", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14)),
      list(x = 0.50, y = 1.02, text = "Type 2", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14)),
      list(x = 0.83, y = 1.02, text = "Other", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14))
    ),
    xaxis = list(title = " "),  # Set shared x-axis title
    yaxis = list(title = " ")   # Set shared y-axis title
  )

facet_plot_combined

```
## Table

```{r}
missing_dt <- data %>%
  group_by(diabetes_type) %>%
  miss_var_summary() %>%
  mutate(pct_miss = round(as.numeric(pct_miss), 2),
         n_miss = format(n_miss, big.mark = ",")) # Round to 2 decimal places and keep as numeric

datatable(missing_dt, extensions = "Buttons", 
            options = list(paging = TRUE,
                           scrollX=TRUE, 
                           searching = TRUE,
                           ordering = TRUE,
                           dom = 'Bfrtip',
                           buttons = c('copy', 'csv', 'excel', 'pdf'),
                           pageLength=5, 
                           lengthMenu=c(3,5,10) ))

```

::::

::: {.callout-tip}

# Summaries


- Core demographic variables such as `age`, `gender`, and `imd_quintile` have no missingness across all diabetes types.

- High missingness is observed primarily in clinical variables related to cholesterol, BMI, statin, and systolic/diastolic values.

- **Type 2 diabetes** has the highest number of records but relatively lower percentages of missingness compared to `Other` and `Type 1`.


:::

## Missingness in Variables Across Patient IDs

This plot visualises the missing data across patient records for all variables in the dataset.

:::: panel-tabset

# Matrix
```{r}
# Missingness across patient id
missing_dt <- data %>% 
  slice_sample(n = 1000) %>% 
  mutate(across(everything(), as.character)) %>% 
  mutate(patient_id = row_number()) %>% 
  pivot_longer(cols = -patient_id, names_to = "variable", values_to = "value") %>% 
  mutate(missing = ifelse(is.na(value), 1, 0)) %>% 
  select(patient_id, variable, missing)


plot_ly(
  data = missing_dt,
  x = ~patient_id,
  y = ~variable,
  z= ~missing,
  type = "heatmap",
  showscale = FALSE
) %>%
  layout(
    title = "Missing Data in Variables Across Patients",
    xaxis = list(title = "Patient ID (not shown) ", showticklabels = FALSE),
    yaxis = list(title = " ")
  )

```
# Table

```{r}

missing_dt <- data %>% 
  slice_sample(n = 1000) %>% 
  mutate(across(everything(), as.character)) %>% 
  mutate(patient_id = row_number()) %>% 
  pivot_longer(cols = -patient_id, names_to = "variable", values_to = "value") %>% 
  mutate(missing = ifelse(is.na(value), 1, 0)) %>% 
  select(patient_id, variable, missing)

datatable(missing_dt, extensions = "Buttons", 
            options = list(paging = TRUE,
                           scrollX=TRUE, 
                           searching = TRUE,
                           ordering = TRUE,
                           dom = 'Bfrtip',
                           buttons = c('copy', 'csv', 'excel', 'pdf'),
                           pageLength=5, 
                           lengthMenu=c(3,5,10) ))

```
::::


::: {.callout-tip}

# Summaries


- Missingness appears non-random for certain variables (e.g., clusters of missing values for clinical variables such as cholesterol-related ones). This is because some variables, such as `cholesterol_category` and `hba1c_mmol_category`, are derived variables; therefore, the distribution of missingness aligns with that of the primary variables from which they are created. However, other variables suggest that specific subgroups or conditions may systematically lack data.

- Variables such as `statin_flag`, `clean_cholesterol_value`, `cholesterol_category`, and `bmi_category` show frequent and widespread 


:::

## Intersections of Missingness Amongst Variables

This is an UpSet plot, which visualises the patterns of missingness, specically the combinations of missingness among variables. 

The horizontal bars on the left show the number of missing values in each variable. In contrast, the vertical bars represent the size of the missing values shared among variables.

The dots and connecting lines beneath the vertical bars indicate which variables' missing values overlap.

:::: panel-tabset

# Plot 

```{r}

gg_miss_upset(data)

```

::::

::: {.callout-tip}

# Summaries

- The horizontal bars on the left show that the variable `clean_cholesterol_value` has the largest number of missing values across all variables. Its derived variable, `cholesterol_category`, follows the same pattern of missingness. This is followed by other variables such as `statin_flag`, `clean_bmi_value`, and the corresponding derived variable, `bmi_category`.

- The largest vertical bar (~22,545 cases) represents the intersection of missing values shared between `clean_cholesterol_value` and its derived variable, `cholesterol_category`.

- There are 4,826 cases where all 5 variables share missing values.

- Intersections involving multiple variables have much smaller sizes compared to single-variable missing values. 


:::
